Compute Servers for Teaching Big Data

### 2025-01-28 paper.tex
data section

The class is called `Big Data'.
For this term to have any meaning, we need to define what we mean by `big'.
For the purposes of this class, a `big' data set is one that is large enough that one cannot load the entire data set into a laptop's memory at once.
Once a data set is this large, it's not possible to simply load the whole thing and start working using R or Python.
Our goal is to choose data sets that are comfortably over this threshold, such that the techniques we're learning are motivated, but not gratuitously large such that answering a basic question takes many hours.
Practically speaking, a typical student laptop will have between 4 and 32 GB of memory at the time of this writing, and data sets that are on the order of 10-100 GB on disk work well to motivate the techniques.
We assign common data analysis tasks that typically take 1 to 10 minutes if done in serial, and will be an order of magnitude faster if done in parallel.

While teaching this course, we have to select data sets.
Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
In general, I'm looking for the following characteristics:
\begin{enumerate}
\item \textbf{Publicly Available} data should be available for legal public download.
\item \textbf{Size} large enough to motivate the computational techniques we're learning, but no larger.
\item \textbf{Format} common format, at least one tabular and another nested. Typically we will do a CSV/TSV for the tabular, and XML or JSON for the nested.
\item \textbf{Background Knowledge} a college student should be able to understand what the data actually represents with minimal effort.
\end{enumerate}

Most of these principles We've violated at one time or another, and the difficulties became clear later.

Amazon Open Data \url{https://registry.opendata.aws/} has provided a reliably good source of large data sets fitting these criterion. 
Another good source is the US government \url{https://data.gov/}.

Once we find a data set to work with, we download it to the server and massage it into the form where we want our students to see.
A large part of the course is `cleaning' the data, preparing it for analysis by filtering, imputing, etc.
TODO: cite
However, it requires judgement as to which cleaning tasks are good for the students learning, and which are merely tedious or too uncommon.

>>>>

%The term `Big Data' appears in the title for Stat 129; for this term to be more than a buzz word, we need to define it.
% TODO: Could talk about how others use this term.
For the purposes of Stat 129, `big data' means a data set that is large enough such that one cannot load the entire data set into a laptop's memory at once.
In a typical introductory data analysis class, a student reads a single data file into R or Python and proceeds with the analysis steps \cite{dsbox}.
This works great if the data fits into memory, but if the data does not fit into memory, then it simply doesn't work.
Data sets that are larger than memory require entirely different approaches, such as streaming through to filter out a subset of interest.
%Once a data set is this large, it's not possible to simply load the whole thing and start working using R or Python.
Our goal is to choose data sets that are comfortably over this memory threshold, such that the techniques we're learning are motivated, but not gratuitously large such that answering a basic question takes an excessive amount of time.
Practically speaking, at the time of this writing, a typical student laptop will have between 4 and 32 GB of memory, and data sets that are on the order of 10-100 GB on disk work well to motivate the techniques.
We assign common data analysis tasks that typically take 1 to 10 minutes if done in serial, and are an order of magnitude faster if done in parallel.

%While teaching this course, we have to select data sets.
%Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
When teaching Stat 129, we look for data sets that satisfy the criteria below.
\begin{enumerate}
\item \textbf{Publicly Available} The data should be available for legal public download.
	Special permissions, non disclosure agreements, and licensing all take time away from teaching.
\item \textbf{Size} The files should be large enough to motivate the computational techniques we're learning, but no larger.
\item \textbf{Format} The storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
	Avoid proprietary, novel, or highly specialized data formats.
	In one semester of Stat 129, we work with data in a plain text tabular format such as CSV or TSV, and nested formats such as XML or JSON.
\item \textbf{Background Knowledge} A college student should be able to understand what the data actually represents.
	If you cannot explain the essential structure and idea in a single class period, then that data is probably too complex.
\end{enumerate}
These criteria don't necessarily apply to more domain specific classes, for example, a bioinformatics class studying genomes would do well to focus on genomic data.
%Most of these principles We've violated at one time or another, and the difficulties became clear later.
Amazon Open Data \url{https://registry.opendata.aws/} has provided a reliably good source of large data sets fitting these criteria. 
Another good source is the US government \url{https://data.gov/}.

We often find interesting data sets that are close, but not quite what we're looking for.
In this case, it may be possible to process the data set before providing it to students to make it better suit the goals of the course.
As a specific example, consider the tax return data on nonprofits provided by the Internal Revenue Service of the United States\cite{irs990}.
We removed the XML namespace information in these files so that students learning XPATH can write \texttt{"//ActivityOrMissionDesc"} to find the `activity or mission description' for a particular nonprofit rather than the cumbersome \texttt{//*[local-name()='ActivityOrMissionDesc']}.
The latter is distracting for students just learning XPATH.

The large data sets are the heart of Stat 129.
To properly analyze them we need to provide our students with a compute server.
